---
title: 'EnWAS on Systolic BP - all cycles separately'
fig-width: 10
fig-height: 8
fig-format: svg
---

# General procedure

Our broad goal is the following:

* Choose an outcome variable: In this case, systolic blood pressure.

* Choose a set of phenotypes to see which, if any, are associated with
  the outcome. For NHANES, a natural choice is the set of nutrition
  variables.

We also know that there are a number of physical and other
characteristics that might be confounders. This may depend on the
choice of outcome and phenotypes. For example, the following are known
to associate with both difference in blood pressure and with choices
in foods.

* Age: Increases with age

* Sex: Men tend to have higher BP until women pass menopause

* Race: Black/African American individuals often have higher BP

* BMI: Strong positive association

* Socioeconomic Status (SES): Lower SES is linked with higher BP

* Education level: Lower education is associated with higher BP

In this example, we are interested in finding NUTRITION PHENOTYPES
that associate with systolic blood pressure. Some choices we make for
the analysis:

A detailed analysis for one cycle is available
[here](EnWAS-SysBP-single-cycle.html). In this document, we follow up
on that analysis by fixing a baseline model and phenotypes of interest
at the very beginning, and replicate the analysis for each cycle
separately.

```{r}
library(nhanesA)
library(phonto)
library(PheWAS)
library(splines)
library(lattice)
library(latticeExtra)
```

# Baseline model and exposure variables

```{r}
base_model <- 'SYSTOLIC ~ RIDAGEYR * RIAGENDR + race + log(BMXBMI) + hbpMeds' # educ + lowincome
exposure_vars <-
    c("SEQN", "DR1DRSTZ", "DRDINT", "DRQSPREP","DR1TNUMF", "DR1TKCAL", "DR1TPROT",
      "DR1TCARB", "DR1TSUGR", "DR1TFIBE", "DR1TTFAT", "DR1TSFAT", "DR1TMFAT",
      "DR1TPFAT", "DR1TCHOL", "DR1TATOC", "DR1TRET"  ,"DR1TVARA", "DR1TBCAR",
      "DR1TCRYP", "DR1TLZ", "DR1TVB1" , "DR1TVB2" , "DR1TNIAC", "DR1TVB6" ,
      "DR1TFOLA", "DR1TFA", "DR1TFF", "DR1TFDFE", "DR1TVB12", "DR1TVC","DR1TVK",
      "DR1TCALC", "DR1TPHOS", "DR1TMAGN", "DR1TIRON", "DR1TZINC", "DR1TCOPP", "DR1TSODI",
      "DR1TPOTA", "DR1TSELE", "DR1TMOIS", "DR1TS040","DR1TS080", "DR1TS120",
      "DR1TS160", "DR1TS180", "DR1TM161", "DR1TM181", "DR1TM201", "DR1TP183",
      "DR1TP204")
```

# Utility functions

```{r}
summarize_var <- function(var, base_model, data, trans = c("none", "asinh", "invnorm"))
{
    trans <- match.arg(trans)
    var <- switch(trans,
                  none = var,
                  asinh = sprintf("asinh(%s)", var), # some have 0 values
                  invnorm = sprintf("invNorm(%s)", var))
    fm <- lm(paste0(base_model, " + ", var), data = data)
    coef(summary(fm))[var, ]
}
add_rownames <- function(d, var = "rownames")
{
    d[[var]] <- rownames(d)
    d
}
perform_enwas <- function(CYCLE = "")
{
    cols <- list(DEMO = c("RIDAGEYR","RIAGENDR","RIDRETH1","DMDEDUC2","INDFMPIR"), 
                 BPX  = c("BPXSY1", "BPXSY2", "BPXSY3", "BPXSY4"),
                 BPQ  = c("BPQ050A","BPQ040A", "BPQ020","BPQ080","BPQ100D"),
                 BMX  = "BMXBMI")
    names(cols) <- paste0(names(cols), CYCLE)
    base_df <- jointQuery(cols)
    bpsys_all <- data.matrix(base_df[c("BPXSY1", "BPXSY2", "BPXSY3", "BPXSY4")])
    bpsys_all[bpsys_all == 0] <- NA
    base_df$SYSTOLIC <- rowMeans(bpsys_all, na.rm = TRUE)
    base_df$INVSYSTOLIC <- 1 / base_df$SYSTOLIC
    base_df <- subset(base_df,
                      RIDAGEYR > 30 & RIDAGEYR < 80 &
                      is.finite(SYSTOLIC))    
    base_df$race <- factor(base_df$RIDRETH1)
    levels(base_df$race)
    levels(base_df$race) <- c("Hispanic+", "Black", "White", "Hispanic+", "Hispanic+")
    ## TODO: Similar for `DMDEDUC2`, maybe `INDFMPIR`
    base_df$hbpMeds <- base_df$BPQ050A
    base_df <- within(base_df,
    {
        hbpMeds[BPQ020 == "No"] <- "No"
        hbpMeds[BPQ020 == "Don't know"] <- NA
        hbpMeds[hbpMeds == "Don't know"] <- NA
    })
    ## extract relevant subset
    keep_vars <-
        c("SYSTOLIC", "INVSYSTOLIC", "SEQN",
          "RIDAGEYR", "RIAGENDR", "race", ## education/"DMDEDUC2", lowincome/"INDFMPIR", 
          "BMXBMI", "hbpMeds", "BeginYear", "EndYear")
    sub_df <- na.omit(base_df[keep_vars])
    ## get diet data
    diet_table_name <- paste0("DR1TOT", CYCLE)
    diet_data <- as.data.frame(nhanes(diet_table_name))
    diet_data <- diet_data[exposure_vars]
    row.names(diet_data) <- diet_data$SEQN
    rownames(sub_df) <- sub_df$SEQN
    common_rows <- intersect(rownames(diet_data), rownames(sub_df))
    df_merged <-
        cbind(sub_df[common_rows, ], diet_data[common_rows, ]) |>
    subset(DR1DRSTZ == "Reliable and met the minimum criteria")
    df_merged <- na.omit(df_merged)
    cat("Number of observations: ", nrow(df_merged), fill = TRUE)
    jointDF <- df_merged
    EVs <- exposure_vars[c(-1,-2,-3,-4)]
    ## perform EnWAS 
    sapply(EVs, summarize_var, base_model = base_model,
           data = df_merged, trans = "invnorm") |>
               t() |> as.data.frame() |> add_rownames("EV") |>
               dplyr::mutate(LCL = `Estimate` - 2 * `Std. Error`,
                             UCL = `Estimate` + 2 * `Std. Error`,
                             adj_pval = p.adjust(`Pr(>|t|)`, "BH"))
}
```

# Per cycle reports

`DR1TOT` is only available from cycle C onwards.

```{r}
nhanesSearchTableNames("DR1TOT")
```


## Cycle C


```{r}
cycle <- "_C"
e <- perform_enwas(cycle)
dotplot(reorder(EV, `t value`) ~ `t value`, data = e, pch = 16,
        groups = ifelse(adj_pval < 0.1, "significant", "not significant"))
segplot(reorder(EV, `t value`) ~ LCL + UCL, data = e, level = Estimate)
topEV <- sort_by(e, ~ adj_pval) |> subset(adj_pval < 0.1) |> getElement("EV")
(lapply(topEV, metadata_var, table = paste0("DR1TOT", cycle)) |>
    do.call(what = rbind))[1:3] |> knitr::kable()
```

## Cycle D

```{r}
cycle <- "_D"
e <- perform_enwas(cycle)
dotplot(reorder(EV, `t value`) ~ `t value`, data = e, pch = 16,
        groups = ifelse(adj_pval < 0.1, "significant", "not significant"))
segplot(reorder(EV, `t value`) ~ LCL + UCL, data = e, level = Estimate)
topEV <- sort_by(e, ~ adj_pval) |> subset(adj_pval < 0.1) |> getElement("EV")
(lapply(topEV, metadata_var, table = paste0("DR1TOT", cycle)) |>
    do.call(what = rbind))[1:3] |> knitr::kable()
```


## Cycle E

```{r}
cycle <- "_E"
e <- perform_enwas(cycle)
dotplot(reorder(EV, `t value`) ~ `t value`, data = e, pch = 16,
        groups = ifelse(adj_pval < 0.1, "significant", "not significant"))
segplot(reorder(EV, `t value`) ~ LCL + UCL, data = e, level = Estimate)
topEV <- sort_by(e, ~ adj_pval) |> subset(adj_pval < 0.1) |> getElement("EV")
(lapply(topEV, metadata_var, table = paste0("DR1TOT", cycle)) |>
    do.call(what = rbind))[1:3] |> knitr::kable()
```


## Cycle F

```{r}
cycle <- "_F"
e <- perform_enwas(cycle)
dotplot(reorder(EV, `t value`) ~ `t value`, data = e, pch = 16,
        groups = ifelse(adj_pval < 0.1, "significant", "not significant"))
segplot(reorder(EV, `t value`) ~ LCL + UCL, data = e, level = Estimate)
topEV <- sort_by(e, ~ adj_pval) |> subset(adj_pval < 0.1) |> getElement("EV")
(lapply(topEV, metadata_var, table = paste0("DR1TOT", cycle)) |>
    do.call(what = rbind))[1:3] |> knitr::kable()
```


## Cycle G

```{r}
cycle <- "_G"
e <- perform_enwas(cycle)
dotplot(reorder(EV, `t value`) ~ `t value`, data = e, pch = 16,
        groups = ifelse(adj_pval < 0.1, "significant", "not significant"))
segplot(reorder(EV, `t value`) ~ LCL + UCL, data = e, level = Estimate)
topEV <- sort_by(e, ~ adj_pval) |> subset(adj_pval < 0.1) |> getElement("EV")
(lapply(topEV, metadata_var, table = paste0("DR1TOT", cycle)) |>
    do.call(what = rbind))[1:3] |> knitr::kable()
```


## Cycle H

```{r}
cycle <- "_H"
e <- perform_enwas(cycle)
dotplot(reorder(EV, `t value`) ~ `t value`, data = e, pch = 16,
        groups = ifelse(adj_pval < 0.1, "significant", "not significant"))
segplot(reorder(EV, `t value`) ~ LCL + UCL, data = e, level = Estimate)
topEV <- sort_by(e, ~ adj_pval) |> subset(adj_pval < 0.1) |> getElement("EV")
(lapply(topEV, metadata_var, table = paste0("DR1TOT", cycle)) |>
    do.call(what = rbind))[1:3] |> knitr::kable()
```


## Cycle I

```{r}
cycle <- "_I"
e <- perform_enwas(cycle)
dotplot(reorder(EV, `t value`) ~ `t value`, data = e, pch = 16,
        groups = ifelse(adj_pval < 0.1, "significant", "not significant"))
segplot(reorder(EV, `t value`) ~ LCL + UCL, data = e, level = Estimate)
topEV <- sort_by(e, ~ adj_pval) |> subset(adj_pval < 0.1) |> getElement("EV")
(lapply(topEV, metadata_var, table = paste0("DR1TOT", cycle)) |>
    do.call(what = rbind))[1:3] |> knitr::kable()
```


## Cycle J

```{r}
cycle <- "_J"
e <- perform_enwas(cycle)
dotplot(reorder(EV, `t value`) ~ `t value`, data = e, pch = 16,
        groups = ifelse(adj_pval < 0.1, "significant", "not significant"))
segplot(reorder(EV, `t value`) ~ LCL + UCL, data = e, level = Estimate)
topEV <- sort_by(e, ~ adj_pval) |> subset(adj_pval < 0.1) |> getElement("EV")
(lapply(topEV, metadata_var, table = paste0("DR1TOT", cycle)) |>
    do.call(what = rbind))[1:3] |> knitr::kable()
```


